---
title: Hopkins package notes
---

Wright, Kevin (2023).
Will the Real Hopkins Statistic Please Stand Up?
The R Journal, 14, 281-291.
https://journal.r-project.org/articles/RJ-2022-055/
https://doi.org/10.32614/RJ-2022-055 (not yet active)

Notes on public literature and code related to Hopkins statistic.

"fixme" means I should contact the authors to suggest corrections.


### R / factoextra

Update: Submitted bug report https://github.com/kassambara/factoextra/issues/160

These two examples are from the `factoextra::get_clust_tendency` function, which uses a copy of the `clustertend::hopkins` code and therefore (incorrectly) uses d=1.


```{r}
library(factoextra)
get_clust_tendency
get_clust_tendency(iris[,-5], n = 50)
# use hopkins( d=1 ) to match
hopkins::hopkins(iris[,-5], m=50, d=1)

# Random uniformly distributed dataset (no inherent clusters)
set.seed(123)
random_df <- apply(iris[, -5], 2, 
                   function(x){runif(length(x), min(x), max(x))}
                   )
get_clust_tendency(random_df, n = 50)
hopkins::hopkins(random_df, m=50, d=1)
```

### R / spatstat.core

Seems to be correct for 2-column data. Probably not for 3-column data etc. fixme

The code for `hopskel` is abstracted, so hard to read, but it appears to use d=2.  Also, the function calculates `A=mean(dX^2)/mean(dU^2)`.  We can convert to the `hopkins::hopkins` scale via 1/(1+A).

Help page says: "The function hopskel calculates the Hopkins-Skellam test statistic A, and returns its numeric value. This can be used as a simple summary of spatial pattern: the value H=1 is consistent with Complete Spatial Randomness, while values H < 1 are consistent with spatial clustering, and values H > 1 are consistent with spatial regularity."

```{r}
pacman::p_load(spatstat.core)
# https://rdrr.io/cran/spatstat.core/src/R/hopskel.R
# https://rdrr.io/cran/spatstat.core/man/hopskel.html
plot(redwood)
hopskel(redwood) # .186 to .223 indicates strong spatial clustering
hopskel.test(redwood, alternative="clustered") # p-value .00001

# Using .20 from hopskel, we can re-scale it to 1/(1+.20) = 0.83,
# which is close to what the hopkins package gives
hopkins::hopkins( as.data.frame(redwood), m=10) # .780 to .835
```

### R / parameters

Update: The code has been re-factored and I cannot find Hopkins statistic.

The code seems to be taken from clustertend::hopkins and likely uses d=1.
fixme

```{r}
# https://github.com/easystats/parameters/blob/main/R/check_clusterstructure.R

library(parameters)
check_clusterstructure(iris[, 1:4])
```

### Python / romusters

Update: submitted bug: https://github.com/romusters/hopkins/issues/1
Appears to use d=1.
https://github.com/romusters/hopkins
https://github.com/romusters/hopkins/blob/master/main.py

### Shushil Deore

Update: Send email to sushildeore99@gmail.com

Appears to use d=1.
https://sushildeore99.medium.com/really-what-is-hopkins-statistic-bad1265df4b

### Python / Hopkins-Test

Update: Submitted bug https://github.com/mhatim99/Hopkins-Test/issues/1

Hard to read.  Appears to use d=1.
https://natureecoevocommunity.nature.com/posts/do-we-do-the-vegetation-data-clustering-analysis-right
https://github.com/mhatim99/Hopkins-Test

### Python / antoniolopezmc

Update: Added comment to this gist

https://gist.github.com/antoniolopezmc/3acd3007a2649d099be4b2e0674d13df

### Matevzkunaver

fixme

Appears to use d=1.
https://matevzkunaver.wordpress.com/2017/06/20/hopkins-test-for-cluster-tendency/

### StackExchange

Update: Added link to publication.

Appears to use d=1.
https://datascience.stackexchange.com/questions/14142/cluster-tendency-using-hopkins-statistic-implementation-in-python?newreg=cea4e39a72e3476f86916513447cb968

### Stackexchange

Update: Added link to publication

The quoted formula from Wikipedia uses "d", but nobody notices that the functions are wrong.
https://stats.stackexchange.com/questions/332651/validating-cluster-tendency-using-hopkins-statistic


### Hopkins-Statistic-Clustering-Tendency

Update: Added link to publication

Appears to use d=1.
https://github.com/prathmachowksey/Hopkins-Statistic-Clustering-Tendency


# Matlab / Fricke

Update: Sent an email with link to publication

Matlab code that uses d=2. `distances(j) = d(u,S(j,:))^2`
https://fricke.co.uk/DandA/hopkins.htm


# Discussion boards etc

PreCLAS: An Evolutionary Tool for Unsupervised Feature Selection.
Used clustertend.  fixme
https://link.springer.com/chapter/10.1007/978-3-030-61705-9_15


Interesting discussion of Hopkins and Holgate indexes.  I cannot find any code for calculating the Holgate index, so this could be an interesting addition. fixme
https://oakmissouri.org/nrbiometrics/
https://oakmissouri.org/nrbiometrics/topics/hopkins.php
https://oakmissouri.org/nrbiometrics/topics/holgate.php



These people notice there's a problem with "d".
https://www.reddit.com/r/AskStatistics/comments/l74bsh/conflicting_definitions_of_hopkins_statistic/


Original post asks "Is there any history of R packages causing wrong results?".  fixme
https://www.reddit.com/r/RStudio/comments/7dcyku/can_r_be_trusted/
Contact Adolfsson and explain error with hopkins().  Post on reddit with possible problem.

Uses factoextra.  fixme
https://mobilemonitoringsolutions.com/us-arrests-hierarchical-clustering-using-diana-and-agnes/

    
    
# Literature reviewed

@adolfsson2019cluster compared different methods to detect clusterability, including the use of `clustertend::hopkins()`


Banerjee. Validating clusters using the Hopkins statistic. Used d=d, H~Beta(m,m) regardless of d.


Besag "on the detection of spatial pattern in plant communities".  Could not find this paper.


Cross 1980. Some approaches to measuring clustering tendency.
Technical Report TR-80–03, Computer Science Dept., Michigan State University, 1980.


Cross and Jain 1982, Measurement of Clustering Tendency,"  Proceedings 
IFAC Symposium on Digital Control, New Delhi,  India,  24-29. 
They extended Hopkins to d dimensions


Diggle 1976. "U is the SQUARED DISTANCE".


Dubes & Jain 1979. Validity studies in clustering methodologies.


Dubes & Zeng. A test for spatial homogeneity in cluster analysis. J Classif 1988 4 33
They say H ~ Beta(m,n). For small d, 5% sampling is suggested. They mention that Cross & Jain 1982 extended Hopkins to d dimensions.


Eberhardt 1967.  Some Developments in 'Distance Sampling'. Biometrics, 23(2), 207–216.


Fernández Pierna 2000. Improved algorithm for clustering tendency. Used d=1. Wheat NIR data 100 obj, 700 variates.


Fowlkes 1988, Variable selection in clustering, J Classification.
Note: This does NOT have car data!


Fowlkes EB, Gnanadesikan R, Kettenring JR.
Variable selection in clustering and other contexts. In C. Mallows (editor), Design, Data, and Analysis, Wiley, New York: 1987;13-34.
Note: Cannot find this book.


Hoffman. A test of randomness based on the minimal spanning tree.


Holgate, P. [1965a]. Some new tests of randomness. J. Ecol. 53, 261-6. 


Holgate, P. [1965b]. Tests of randomness based on distance methods. Biometrika 52, 345-53. 


Brian Hopkins (1957). The Concept of Minimal Area. Journal of Ecology, 45(2), 441–449.   doi:10.2307/2256927


Jurs, Peter C.; Lawson, Richard G. (1990). Clustering Tendency Applied to Chemical Feature Selection. Drug Information Journal, 24(4), 691–704. doi:10.1177/216847909002400405  
They claimed Fowlkes (Variable selection in clustering and other contexts) used cars data 56x9: price, mpg, seat len, trunk size, weight, length, radius, displacement, gear ratio.  Note, I cannot find this data!


Kittler. Pattern Recognition Theory and Applications.
https://www.google.com/books/edition/Pattern_Recognition_Theory_and_Applicati/_QerCAAAQBAJ?hl=en&gbpv=1&dq=hopkins+statistic&pg=PA93&printsec=frontcover
Uses d=d.


Lawson and Peter C. Jurs. New index for clustering tendency and its application to chemical problems. Journal of Chemical Information and Computer Sciences , 30(1):36–41, 1990.  Used d=1. Suggest 10% sampling of points, but did also use higher fractions. Note: Figure 4 (page 39) is a 2-d scatterplot with calculated Hopkins values in table VI. Could be useful to verify they used d=1 to calculate Hopkins.


Li, Fasheng; Lianjun Zhang 2007.
Comparison of point pattern analysis methods for classifying the spatial distributions of spruce-fir stands in the north-east USA
Forestry: An International Journal of Forest Research, Volume 80, Issue 3, July 2007, Pages 337–349, https://doi.org/10.1093/forestry/cpm010
" The toroidal edge correction proved to be simple and satisfactory in this study compared with other edge correction methods."


Erdal Panayirci; Richard C Dubes (1983). A test for multidimensional clustering tendency. Pattern Recognition, 16(4), 433–444. doi:10.1016/0031-3203(83)90066-3
Uses d=d.


Petrie, Adam; Willemain, Thomas R. (2013). An empirical study of tests for uniformity in multidimensional data. Computational Statistics & Data Analysis, 64(), 253–268. doi:10.1016/j.csda.2013.02.013
Fairly recent study. Tests based on minimum spanning tree and "snake run length" are proposed and evaluated for multidimensional data.


E. C. Pielou. The Use of Point-to-Plant Distances in the Study of the Pattern of Plant Populations


Pielou 1960. A Single Mechanism to Account for Regular, Random and Aggregated Populations. Not useful.


Rotondi, Renata (1993). Tests of Randomness Based on the K-NN Distances for Data from a Bounded Region. Probability in the Engineering and Informational Sciences, 7(4), 557–569. doi:10.1017/S0269964800003132


Smith, Stephen. 1982. Structure of Multidimensional Patterns.
https://d.lib.msu.edu/etd/36982


Smith, Stephen. 1984. Testing for uniformity in multidimensional data.
Uses minimum spanning tree which assumes that the convex hull is a reasonable estimate of the sampling window.


Varmuza. Introduction to Multivariate Statistical Analysis in Chemometrics.
https://www.google.com/books/edition/Introduction_to_Multivariate_Statistical/S4btpoIgGP0C?hl=en&gbpv=1&dq=%22hopkins+statistic%22&pg=PA272&printsec=frontcover
Uses d=1.


Yang 2017. Multivariate tests of uniformity.

Zeng & Dubes, A Comparison of tests for randomness.
